---
title: "01-benefit-per-person"
author: "Juliane Kloidt"
date: "2025-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# set the workspace path to the directory containing this script
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# clear the environment
rm(list=ls())

library(tidyverse)

data <- read_csv("data/data-scotland.csv") %>%
  select(-c(damage_pathway, damage_type, households, nation))

```


```{r}
# combine co-benefits for each factor, grouped by local authority, and small area
data_benefit_factor <- data %>% group_by(local_authority, small_area, `co-benefit_type`, population) %>%
  summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)))

# combine co-benefits across all factors, grouped by local authority, and small area
data_benefit_sum <- data %>% group_by(local_authority, small_area, population) %>%
  summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)))

## missing data for small_area == S01010206 and `co-benefit_type` == hassle_costs

# combine co-benefits across all factors and across small areas, grouped by local authority
data_benefit_sum_authority <- data %>% group_by(local_authority, small_area, population) %>%
  summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>% ungroup() %>%
  group_by(local_authority) %>% summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)))

# sum of co-benefits for each local authority
ggplot(data_benefit_sum_authority, aes(y=sum, x = local_authority)) +
  geom_col() + ylab("Benefits per local authority (Million GBP)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

# 
top_five <- c("physical_activity", "air_quality", "noise", "diet_change", "excess_cold")

data_benefit_local <- data %>% rowwise() %>% mutate(weighted = sum/population) %>%
  group_by(local_authority, `co-benefit_type`) %>%
  summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)))
  filter(`co-benefit_type` %in% top_five)

ggplot(data_benefit_local, aes(y=weighted, x=local_authority, fill=`co-benefit_type`)) +
  geom_bar(stat = "identity", position="stack") +
  ylab("Benefits per local authority (Million GBP)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

data_benefit_local %>%
  filter(`co-benefit_type` == "physical_activity") %>%
  arrange(desc(weighted)) %>%
  ggplot(., aes(x=local_authority, y=weighted)) +
  geom_bar(stat="identity") +
  ylab("Benefits per local authority (Million GBP)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())
```

# Research Question 1: How much does each Scottish local authority benefit from net zero?

Which authorities benefit the most from reaching net zero? Are there meaningful differences between groups of local authorities (e.g., urban vs. rural)?

```{r}
# sum of co-benefits for each local authority weighted by population
data_benefit_sum_authority %>%
  mutate(sum_weighted = (sum/population)*1000000) %>%
  ggplot(., aes(y=sum_weighted, x = local_authority)) +
  theme_classic()+
  geom_col() + ylab("Benefits per person (GBP)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

# distribution
data_benefit_sum_authority %>%
  mutate(sum_weighted = (sum/population)*1000000) %>%
  ggplot(., aes(x=sum_weighted)) +
  geom_density()

# summary statistics
data_benefit_sum_authority %>%
  mutate(sum_weighted = (sum/population)*1000000) %>%
  select(local_authority, sum_weighted) %>%
  summarise(mean = mean(sum_weighted),
            median = median(sum_weighted),
            sd = sd(sum_weighted),
            min = min(sum_weighted),
            max= max(sum_weighted))

# rank order authorities by co-benefit value
data_benefit_sum_authority %>%
  mutate(sum_weighted = (sum/population)*1000000) %>%
  select(local_authority, sum_weighted) %>%
  arrange(desc(sum_weighted))

# top 5 beneficiaries: Angus, East Renfrewshire, East Dunbartonshire, Falkirk, Orkney Islands
# bottom 5 beneficiaries: Glasgow City, Moray, Scottish Borders, Dumfries and Galloway, East Ayrshire
```

# Research Question 2: When does each local authority benefit from net zero?

Which authorities benefit at what time on the path to net zero? Which authorities have an early pay-out, late pay-out, constant pay-out? Are there meaningful differences between groups of local authorities (e.g., urban vs. rural)?

Do authorities need to invest up front, or do they get a pay out?

In what year does the local authority receive 25%, 50%, and 75% of the pay out per person?

```{r}
# plot trajectory of benefits for each local authority WEIGHTED BY POPULATION
data_benefit_authority_long_weighted <- data_benefit_sum_authority %>% 
  rowwise() %>% mutate(across(c(`2025`:`2050`), .fns = ~./population)) %>%
  select(-c(population, sum)) %>%
  pivot_longer(cols = -c(local_authority), names_to = "year", values_to = "benefits") %>%
    mutate(benefits = benefits*1000000)

ggplot(data_benefit_authority_long_weighted, 
       aes(x=year, y=benefits, group = local_authority, color = local_authority)) +
  theme_minimal() +
  geom_line()
# there seems to be several groups of authorities: 
# some increase strongly and then decrease, others increase more continuously
# 0.00010 billion is 0.1 million is 100K benefit per person

# try to spot patterns in which local authorities have a spike in benefits, and which have a gradual increase
# rural / islands show spike > they would immediately benefit from net zero
# cities (Glasgow, Edinburgh, Dundee, Aberdeen, are very similar)
ggplot(data_benefit_authority_long_weighted, 
       aes(x=year, y=benefits, group = local_authority, color = local_authority)) +
  theme_minimal() +
  geom_line() +
  facet_wrap(~local_authority) +
  theme(legend.position = "none",
        axis.text.x = element_blank())

# questions: are there distinct groups of local authorities (rural vs urban)
# OR does the trajectory of co-benefits depend mostly on population (e.g., 100K+ vs. 100K-) > Stirling is under 100K
# OR does the trajectory of co-benefits depend on geographical location (e.g., north vs south)


```