---
title: "02-figures-final"
author: "Juliane Kloidt"
date: "2025-12-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up
```{r}
# set the workspace path to the directory containing this script
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# clear the environment
rm(list=ls())

library(tidyverse)
library(sf)
library(viridis)
library(patchwork)
library(ggmosaic)

# co-benefit data (Scotland only)
data3 <- read_csv("data/data3-scotland.csv")

# SIMD data (plus spatial data for Glasgow City)
Ggrid <- st_read("data/SG_SIMD_2020.shx")
Ggrid_glasgow <- Ggrid %>% filter(LAName == "Glasgow City")
```

# The Urban Dividend Gap: Unmasking Inequity in Scotland’s Net Zero Transition

## 1. The Urban Burden: Major Cities benefit less from Net Zero but pay similar Hassle Tax

```{r fig 1, level 1 data}
net_benefit <- data3 %>%
  select(small_area, local_authority, population, sum) %>%
  group_by(local_authority) %>%
  summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
  mutate(net_benefit = (sum/population)*1000000) %>%
  select(-c(sum, population)) %>%
  ungroup()

gross_benefit <- data3 %>%
  mutate(sum_gross = select(., c(air_quality, dampness:excess_heat, 
                                     noise: physical_activity)) %>% rowSums(na.rm = TRUE)) %>%
  select(small_area, local_authority, population, sum_gross) %>%
  group_by(local_authority) %>%
  summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
  mutate(gross_benefit = (sum_gross/population)*1000000) %>%
  select(-c(sum_gross, population)) %>%
  ungroup()

hassle_tax <- inner_join(gross_benefit, net_benefit) %>%
  mutate(hassle_tax = gross_benefit - net_benefit) %>%
  arrange(desc(hassle_tax))

fig1_data <- inner_join(gross_benefit, net_benefit) %>%
  arrange(desc(net_benefit)) %>% rowid_to_column() %>%
  pivot_longer(cols=c(gross_benefit, net_benefit), names_to = "benefit_type", values_to = "value")

fig1 <- ggplot(fig1_data, aes(x=value, y=reorder(local_authority, -rowid)))+
  theme_classic()+
  geom_line(aes(group = local_authority), color="#3e4989")+
  geom_point(aes(colour = benefit_type))+
  labs(x = "Co-Benefits per Capita (GBP)", y = NULL,
       title = "Gross Potential vs. Net Gain per Capita across Scottish Local Authorities, 2025 to 2050")+
  scale_color_manual(labels = c("Gross Potential = Benefits", 
                                "Net Gain = Benefits - Costs"), 
                     values = c("#939184", "#ed4a89"))+
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.8, 0.15),
        plot.title.position =  "plot" ,
        plot.caption.position =  "plot",
        text = element_text(color = "#3b3a34"),
        plot.caption = element_text(hjust = 0, color = "#706e63"),
        plot.background = element_rect(fill = "#f6f5f4",color = NA),
        panel.background = element_rect(fill = "#f6f5f4",color = NA),
        legend.background = element_rect(fill = "#f6f5f4",color = NA),
        plot.margin = unit(c(.5, .5, .2, .5), "cm"))

fig1

ggsave(filename = "figures/fig1-urban-burden.png", 
       bg="white", width = 210, height = 140, units = "mm")


#######################
# fig1 with caption and title
fig1_caption <- fig1 +
  labs(title = "Major Cities Benefit Less from Net Zero, But Pay Similar Hassle Tax",
       subtitle = "Gross Potential vs. Net Gain per Capita across Scottish Local Authorities, 2025 to 2050",
    caption = "Analysis across Scotland's 32 local authorities reveals that major cities like Glasgow and Edinburgh are positioned at the\nbottom of the 'Net Gain' rankings. These urban centers face lower per-capita co-benefit potential than car-dependent\ncommuter belts like Angus or East Renfrewshire, yet they pay a similar 'Hassle Tax', the economic\nand behavioural friction associated with complex infrastructure changes.")

fig1_caption

ggsave(filename = "figures/fig1-urban-burden-caption.png", 
       bg="white", width = 210, height = 140, units = "mm")
```

## 2. The Composition Gap: Why Cities Lag

```{r}
# present top 3 and bottom 3 beneficiaries
LA <- c("Angus", "East Renfrewshire", "East Dunbartonshire",
        "Glasgow City", "Moray", "Scottish Borders")

top3 <- c("Angus", "East Renfrewshire", "East Dunbartonshire")
bottom3 <- c("Glasgow City", "Moray", "Scottish Borders")

fig2_data <- data3 %>%
  filter(local_authority %in% LA) %>%
  select(local_authority, population, physical_activity, 
         air_quality, noise, excess_cold, diet_change) %>%
  group_by(local_authority) %>%
  summarize(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
  rowwise() %>%
  mutate(activity_weighted = (physical_activity/population)*1000000,
         air_weighted = (air_quality/population)*1000000,
         noise_weighted = (noise/population)*1000000,
         cold_weighted = (excess_cold/population)*1000000,
         diet_weighted = (diet_change/population)*1000000) %>%
  select(-c(population, physical_activity, air_quality, noise, excess_cold, diet_change)) %>%
  pivot_longer(cols = c(activity_weighted, air_weighted, noise_weighted, cold_weighted, diet_weighted),
               names_to = "co_benefit", values_to = "weighted") %>%
  mutate(weighted = round(weighted))%>%
  mutate(divider = case_when(local_authority %in% top3 ~ "Top 3 Local Authorities",
                             local_authority %in% bottom3 ~ "Bottom 3 Local Authorities"))

fig2_data$local_authority <- 
  gsub("East Renfrewshire", "East\nRenfrewshire", fig2_data$local_authority)
fig2_data$local_authority <- 
  gsub("East Dunbartonshire", "East\nDunbartonshire", fig2_data$local_authority)

fig2_data$divider <- factor(fig2_data$divider,
                         levels = c("Top 3 Local Authorities", "Bottom 3 Local Authorities"))



fig2 <-ggplot(fig2_data, aes(y=weighted, x=local_authority, 
                             fill = reorder(co_benefit, weighted),label = weighted)) +
  theme_classic()+
  geom_bar(stat = "identity", position = "stack") +
  geom_text(size = 2.5, position = position_stack(vjust = 0.4), color="white")+
  facet_grid(~divider, scales = "free", space = "free") +
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size=12, face = "bold", color = "#3b3a34"),
        legend.title = element_blank(),
        plot.title.position =  "plot" ,
        plot.caption.position =  "plot",
        text = element_text(color = "#3b3a34"),
        plot.caption = element_text(hjust = 0, color = "#706e63"),
        plot.background = element_rect(fill = "#f6f5f4",color = NA),
        plot.margin = unit(c(.5, .5, .2, .5), "cm"),
        legend.background = element_rect(fill = "#f6f5f4",color = NA),
        panel.background = element_rect(fill = "#f6f5f4",color = NA),
        strip.background = element_rect(fill = "#f6f5f4",color = NA))+
  scale_fill_manual(labels = c("Diet Change", "Excess Cold\nReduction", 
                                "Air Quality", "Noise\nReduction",
                                "Physical\nActivity"), 
                     values = c("#ed6925", "#ed4a89", "#26828e", "#31688e", "#3e4989"))+
  labs(y = NULL,
    title = "Composition of per Capita Gains (GBP) in the Top and Bottom Local Authorities, 2025 to 2050")

fig2

ggsave(filename = "figures/fig2-proportion-benefits.png", bg="white", 
       width = 210, height = 140, units = "mm")

#######################
# fig2 with caption and title

fig2_caption <- fig2 +
  labs(title = "The Composition Gap: Why Cities Lag", 
    subtitle = "Composition of per Capita Gains (GBP) in the Top and Bottom Local Authorities, 2025 to 2050", 
       caption = "The disparity in potential is driven by baseline behaviors. High-gain areas benefit from massive shifts away from car dependency, translating into\nhigh scores for physical activity and air quality. In cities where public transport use is already higher, the 'delta' for improvement is smaller, making\nthe transition high-effort but relatively lower-reward on a per-person basis.")

fig2_caption

ggsave(filename = "figures/fig2-proportion-benefits-caption.png", bg="white", 
       width = 210, height = 120, units = "mm")

```

## 3. The Internal Fracture: Wealth-Concentrated Rewards

```{r}
data_sum <- data3 %>%
  filter(local_authority=="Glasgow City") %>%
  mutate(weighted = (sum/population)*1000000) %>%
  rename("DataZone" = "small_area") %>%
  select(DataZone, population, weighted)

grid_glasgow_sum <- left_join(Ggrid_glasgow, data_sum)

fig3 <- ggplot(grid_glasgow_sum) +
  geom_sf(mapping = aes(fill=weighted), color="white")+
  theme_void()+
  scale_fill_viridis(
    option = "magma",
    name = "Co-Benefits\nper Capita\n(GBP)",
    alpha = 0.8,
    begin = 0.1, 
    end = 0.9,
    discrete = F,
    direction = 1, # dark is lowest, yellow is highest
    guide = guide_colorbar(
     keyheight = unit(30, units = "mm"),
     title.position = "top",
     reverse = F # display highest co-benefits on top
  )) +
  labs(x = NULL, y = NULL,
         title = "Glasgow City's Co-Benefits per Capita across Data Zones, 2025 to 2050")+
  theme(plot.title.position =  "plot" ,
        plot.caption.position =  "plot",
        text = element_text(color = "#3b3a34"),
        plot.caption = element_text(hjust = 0, color = "#706e63"),
        plot.background = element_rect(fill = "#f6f5f4",color = NA),
        plot.margin = unit(c(.5, .5, .2, .5), "cm"),)

fig3

ggsave(filename = "figures/fig3-glasgow-inequal.png", 
       bg="white", width = 180, height = 120, units = "mm")

#######################
# fig3 with caption and title

fig3_caption <- fig3 +
  labs(title = "The Internal Fracture: Wealth-Concentrated Rewards",
         subtitle = "Glasgow City's Co-Benefits per Capita across Data Zones, 2025 to 2050",
       caption = "Zooming into Glasgow, the city with the nation's lowest average net gain, reveals a stark geographical divide.\nMapping the total co-benefits across small areas (Data Zones) shows that the rewards of the transition are not\nreaching deprived communities. Instead, they are disproportionately concentrated in the city’s most affluent\npockets, such as the West End."
       )

fig3_caption

ggsave(filename = "figures/fig3-glasgow-inequal-caption.png", 
       bg="white", width = 180, height = 120, units = "mm")

```

## 4. The Smoking Gun: The Housing Anomaly

```{r}
data_cold <- data3 %>% select(local_authority, small_area, population, excess_cold)%>%
  filter(local_authority=="Glasgow City") %>%
  rowwise() %>%
  mutate(sum_weighted = (excess_cold/population)*1000000) %>%
  group_by(small_area, population) %>%
  summarize(weighted=sum(sum_weighted, na.rm = TRUE)) %>% 
  ungroup() %>%
  rename("DataZone" = "small_area") %>%
  select(DataZone, population, weighted)

grid_glasgow_cold <- left_join(Ggrid_glasgow, data_cold)

fig4 <- ggplot(grid_glasgow_cold, aes(x=Percentv2, y = weighted, color=weighted)) +
  theme_classic()+
  geom_jitter() +
  geom_smooth(method="lm", color="#ed4a89", fill="white") +
  labs(x = "Scottish Index of Multiple Deprivation (SIMD Percentile)", y = "Reduction in Excess Cold",
         title = "Co-Benefit Inequalities: The Housing Anomaly",
         subtitle = "Reducing Excess Cold (2025 to 2050) vs. Multiple Deprivation (2020) across Glasgow's Data Zones",
       caption = "The most damning evidence of this equity gap is found in the housing sector. Statistically, the health-related benefit of reducing\n'Excess Cold' is positively correlated with affluence (SIMD Percentile). This suggests that current retrofit pathways prioritize\nhigh-value, hard-to-treat historic tenement stock in wealthy areas, effectively bypassing those suffering from the most acute\nfuel poverty."
       )+
  theme(plot.title.position =  "plot" ,
        plot.caption.position =  "plot",
        text = element_text(color = "#3b3a34"),
        plot.caption = element_text(hjust = 0, color = "#706e63"),
        plot.background = element_rect(fill = "#f6f5f4",color = NA),
        plot.margin = unit(c(.5, .5, .2, .5), "cm"),
        legend.background = element_rect(fill = "#f6f5f4",color = NA),
        panel.background = element_rect(fill = "#f6f5f4",color = NA))+
  scale_color_viridis(
    option = "magma",
    name = "Gains\nper Capita\n(GBP)",
    alpha = 0.8,
    begin = 0.1, 
    end = 0.9,
    discrete = F,
    direction = 1, # dark is lowest, yellow is highest
    guide = guide_colorbar(
     keyheight = unit(30, units = "mm"),
     title.position = "top",
     reverse = F
  )) 

fig4

ggsave(filename = "figures/fig4-glasgow-inequal-housing.png", bg="white", width = 180, height = 140, units = "mm")
```

## Additional Plots
Make sure the argument for figure 4 holds up when considering related, less influential co-benefits.
```{r}
data_damp <- data %>% filter(`co-benefit_type`=="dampness") %>%
  filter(local_authority=="Glasgow City") %>%
  rowwise() %>%
  mutate(sum_weighted = (sum/population)*1000000) %>%
  group_by(small_area, population) %>%
  summarize(weighted=sum(sum_weighted, na.rm = TRUE)) %>% 
  ungroup() %>%
  rename("DataZone" = "small_area") %>%
  select(DataZone, population, weighted)

grid_glasgow_damp <- left_join(Ggrid_glasgow, data_damp)


ggplot(grid_glasgow_damp, aes(x=Percentv2, y = weighted)) +
  theme_classic()+
  geom_jitter() +
  geom_smooth(method="lm") +
  ylab("Co-Benefits from Reducing Dampness") + xlab("SIMD Percentile")
```

```{r}
data_heat <- data %>% filter(`co-benefit_type`=="excess_heat") %>%
  filter(local_authority=="Glasgow City") %>%
  rowwise() %>%
  mutate(sum_weighted = (sum/population)*1000000) %>%
  group_by(small_area, population) %>%
  summarize(weighted=sum(sum_weighted, na.rm = TRUE)) %>% 
  ungroup() %>%
  rename("DataZone" = "small_area") %>%
  select(DataZone, population, weighted)

grid_glasgow_heat <- left_join(Ggrid_glasgow, data_heat)


ggplot(grid_glasgow_heat, aes(x=Percentv2, y = weighted)) +
  theme_classic()+
  geom_jitter() +
  geom_smooth(method="lm") +
  ylab("Co-Benefits from Reducing Excess Heat") + xlab("SIMD Percentile")
```

